---
title: "Intro to R for Researchers"
author: 
  - name: Sichao Wang
    degrees: PhD
    email: wangsic1@msu.edu
    affiliations: 
      - name: Michigan State University, Center for Statistical Training and Consulting
date: "`r format(Sys.time(), '%Y-%m-%d, %H:%M:%S %Z')`"
title-block-banner: "#01665e"
format: 
  html: 
    toc: true 
    toc-depth: 4
    toc-location: left 
    number-sections: true
    number-depth: 4
    code-tools: false 
    embed-resources: true 
    theme: default 
execute: 
  eval: false # do not render the code, only include the output in the final document
  echo: true #shows code in the final document
  output: true # include the output in the final document
  message: false  #suppress messages in the final document
  warning: false  #suppress warnings in the final document
  error: false  #suppress errors in the final document
  include: true  # include the code and its output in the final document
knitr:
  opts_chunk:
    out.width: "150%"
bibliography: references.bib
editor:
  markdown:
    wrap: sentence
---

# Project title: Intro to R for Researchers

Quarto document is a markdown-based format for authoring documents, presentations, and websites. There are hierarchical sections, which are defined by the number of #. The more #, the lower level of section. You can use different level of sections to organize your document.


```{r}
#| label: load-packages
#| eval: true
#| echo: true

# At the very beginning of each script or R session. You usually load the required packages and set seed for random process. 

library(tidyverse)
library(medicaldata)
library(readr)
library(Hmisc)
library(ggmosaic)
library(GGally)


set.seed(123) # set seed for random process. Any number can be used. I like to use my MSU ZPIN.
```



# R built-in Help facility
## Help functions

```{r}
#| label: help-functions
#| eval: false
#| echo: true


help.start()  # web-based r help pages.

help(mean)    # bring up the help for the specific function in the help panel.

?mean         # alternative way of help(mean)


methods(class="data.frame") #showing generic function for data.frame class

methods(plot) #what class generic function takes

getS3method("print", "data.frame") #method used for data frames
getS3method("print", "lm") #method used for linear model results

plot #  UseMethod("plot") means generic

getAnywhere("plot.acf")

version #check current R version

```


## Demostration

```{r}
#| label: demonstration
#| eval: false
#| echo: true

demo()

demo(package = .packages(all.available = TRUE))

demo(lm.glm)
# demo(notebook,package = "knitr")

help(plotmath)
example(plotmath)
demo(plotmath)
```


# Key concepts

## Objects

### Types of object

1. atomic *vector*

```{r}
#| label: atomic-vector
#| eval: true
#| echo: true

z <- rnorm(100)
```


2. *matrices* or more generally *arrays* are multi-dimensional generalizations of vectors.


```{r}
#| label: matrices-arrays
#| eval: true
#| echo: true
# An array can be considered as a multiply subscripted collection of data entries, for example numeric
dim(z) <- c(2,5,10) #2 by 5 by 10 array
```

3. *factors* provides compact ways to handle categorical data

order or unorder

```{r}
#| label: factors
#| eval: true
#| echo: true
# unordered character vector, c() is canoncatenate function, which can combine values into a vector or list.


fruits <- c("blueberry","strawberry","apple","pineapple","watermelon")
fruits_f <- factor(fruits)


# notice that the default order of levels is alphabetic, which might not be what we want. 
levels(fruits_f)


# We can change the order of levels by assign new levels to the factor object.

levels(fruits_f) <- c("blueberry","strawberry","apple","pineapple","watermelon")

# Ordered

fruits_order <- factor(fruits,ordered = TRUE)




levels(fruits_order) <- c("blueberry","strawberry","apple","pineapple","watermelon")

fruits_order

#removes the "class" attribute from an object. It reveals the underlying code of ordered factor, which is an integer vector with attributes of levels and class.

unclass(fruits_order) 


# convert ordered factored into numeric vector.
as.numeric(fruits_order)
```


4. *lists* are a general form of vector in which the various elements need not be of the same type, and are often themselves vectors or lists.


```{r}
#| label: lists
#| eval: false
#| echo: true


# Put different types of objects into a list. A list can contain any type of object, including other lists. It is a very flexible data structure in R.
list0 <- list(student="Joe", subject="Math", no.exam=3, scores=c(85,89,93))


# indexing a list is different from indexing a vector. For a list, single square brackets [ ] will return a sublist, while double square brackets [[ ]] will return the actual element.


list0[[2]]


list0$subject
list0[["subject"]]

list0[[4]][1]
list0$scores[1]
list0[["scores"]][1]

```


5. *data frames* are matrix-like structures (2 dimension), in which the columns can be of different types. 

A data frame may for many purposes be regarded as a matrix with columns possibly of differing data types. It may be displayed in matrix form, and its rows and columns extracted using matrix indexing conventions

```{r}
#| label: data-frames
#| eval: true
#| echo: true
fruits <- data.frame(names=c("blueberry","apple","pineapple"),
                     color=c("blue","red","pineapple"),
                     weigh_lb=c(0.05,0.5,5))
```


5. *functions* are themselves objects in R which can be stored in the project’s workspace. This provides a simple and convenient way to extend R. 

All objects in R have a class, reported by the function class. For simple vectors this is just the mode, for example "numeric", "logical", "character" or "list", but "matrix", "array", "factor" and "data.frame" are other possible values


### Properties of object

```{r}
#| label: object-properties
#| eval: false
#| echo: true
mode(z) # the type of atomic component

length(z) # how many of the basic component

attr(z, "dim") <- c(10,10) # add dimension attribute to z, which is now a matrix. 

attributes(z)

```

### Demo for object 

```{r}
#| label: object-demo
#| eval: true
#| echo: true
# create a numeric vector of 100 random normal values named x.

x <- rnorm(100) 

# Then use x as input for generate another numeric vector named y, which is also a random normal vector with mean of x.


y <- rnorm(x)

plot(x, y)

# ls() # list all the objects lives in the environment.
rm(x, y) # delete objects from environment.
```



```{r}
#| label: linear-regression-demo
#| eval: false
#| echo: true
x <- 1:20 # create a numeric vector of 20 integers from 1 to 20.
w <- 1 + sqrt(x)/2
dummy <- data.frame(x=x, y= x + rnorm(x)*w)
dummy
fm <- lm(y ~ x, data=dummy) # fit a linear regression model with y as outcome and x as predictor, and save the fitted model as fm.
summary(fm)

fm1 <- lm(y ~ x, data=dummy, weight=1/w^2) # fit a weighted linear regression model with y as outcome and x as predictor, and save the fitted model as fm1.
summary(fm1)
attach(dummy) #Make the columns in the data frame visible as variables.


# scatter plot with lowess curve and regression line.
lrf <- lowess(x, y)
plot(x, y)
lines(x, lrf$y)
abline(0, 1, lty=3)
abline(coef(fm))
abline(coef(fm1), col = "red")
detach() #Remove data frame from the search path.

# Residuals vs. fitted plot

plot(fitted(fm), 
     resid(fm), 
     xlab="Fitted values", 
     ylab="Residuals", 
     main="Residuals vs Fitted")

# QQ plot for residuals

qqnorm(resid(fm), main="Residuals Rankit Plot")
qqline(resid(fm))
rm(fm, fm1, lrf, x, dummy)

```


## Data modes/types

### numeric

```{r}
#| label: numeric-type
#| eval: false
#| echo: true

a<-2
b <- a
class(a)

# integer vector
vec1<-c(1L,2L,3L,4L,5L,6L)

vec2<-c(1L,2L)

class(vec1)

# recyling rule: when two vectors have different length, the shorter one will be recycled to match the length of the longer one. If the length of the longer vector is not a multiple of the shorter one, a warning will be issued.
vec1 + vec1 

vec1 / vec2  # notice that shorter vector will be recycled for operation.

sqrt(vec2)

n <- 10
1:n-1
1:(n-1) # notice the mathematical operation order, () take precedence.

# useful way to generate sequence. pretty useful to create index sequence in for loop, or plot ticks labels.

seq(1,30)
seq(from=1, to=30)
seq(to=30, from=1)

```
There are some common mathematical operator we can apply to numeric, such as  +, -, *, / and ^ . Also some of functions carry mathematical operation, such as log, exp, sin, cos, tan, sqrt.

### Character string

```{r}
#| label: character-string
?Quotes

# use can use double quotes or single quotes to create character string. If you want to use the quote itself, you can use the other type of quote to create the string, or you can use escape character \ to escape the quote.
greet <- "hello world"  

class(greet)

greet2 <- 'hello world'

greet3 <- '"hello world"'

greet4 <- "hello \"world\""

difference of greet3 <- '"hello world"' and greet4 <- "hello \"world\""

```

### logical

```{r}
#| label: logical-type
#| eval: false
#| echo: true

logical1 <- TRUE
logical2 <- FALSE

class(logical2)

# stored as numeric vector with 0 for FALSE and 1 for TRUE, so we can do mathematical operation on logical vector. This is often used in data manipulation, such as subsetting data by condition, or calculating the proportion of missing data.
logical1+logical2


# A logical vector.

logical2 <- c(TRUE, TRUE,FALSE)

#as.something to help you manipulate types.

as.numeric(logical1) #coerced into numeric vectors, FALSE becoming 0 and TRUE becoming 1.


as.character(logical1)
```
There are also data type as integer, complex and raw, which are not very commonly used. Be careful about as.something, it might have unexpected result.


### Missing data

In general, any mathematical or logical operation involving an `NA` results in an `NA`. For this reason, checking for missing values with `x == NA` will not work; you must use `is.na(x)` instead. R also uses `NaN` (Not a Number) for undefined mathematical operations, such as dividing by zero. Note that `is.na(x)` returns `TRUE` for both `NA` and `NaN` values. To specifically identify `NaN` values, use `is.nan(x)`.

```{r}
#| label: missing-data
#| eval: false
#| echo: true


z <- c(1:3,NA); ind <- is.na(z)

z == NA #NA basically an unknown value, so we cannot know if it is equal to NA or not. So it returns NA.
```


## Indexing(subsetting)


```{r}
#| label: indexing
#| eval: false
#| echo: true

x <- rnorm(10)

# by index number
x[1:3] # A vector of positive integral quantities select by position
x[-3]  # A vector of negative integral quantities deselect by position

# by conditional, combine logical operator and indexing to subset data by condition. This is very useful for data manipulation, such as subsetting data by condition, or calculating the proportion of missing data. ! is negate operator.
y <- x[!is.na(x)&x>0] 

# by name
fruit <- c(5, 10, 1, 20) 
names(fruit) <- c("orange", "banana", "apple", "peach") 
lunch <- fruit[c("apple","orange")]

#index help you to manipulate vector such as

x[2] <- 44 # replace an element.
x[11] <- 100 # extend vector. 

```

Some conditional ooperator  <, <=, >, >=, == for exact equality and != for inequality
c1 & c2 is their intersection (“and”), c1 | c2 is their union (“or”), and !c1 is the negation of c1.

## Distribution

In R, distribution functions use a standard naming convention. Prefix the base name of the distribution with **'d'** for the density, **'p'** for the cumulative distribution function (CDF), **'q'** for the quantile function, and **'r'** for simulation (random deviates).

```{r}
#| label: distributions
#| eval: false
#| echo: true

2*pt(-2.43, df = 13) # gives the two-tailed p-value for a t-test with 13 degrees of freedom and a t-statistic of -2.43.
qf(0.01, 2, 7, lower.tail = FALSE) # gives the critical value of the F-distribution with 2 and 7 degrees of freedom at the 1% significance level (upper tail).
```



# IDA: Descriptive data analysis

## Import data

There are a few ways to import/load dataset into your workspace.

1. load dataset from your local directory.
2. load dataset from online depository.
3. load dataset from package.
4. connect to database and get data


```{r}
#| label: import-data
#| eval: true
#| echo: true

# 1. load dataset from your local directory.
# Depends on the data format, you might need different function from different package to load data. Common package readxl for xls and xlsx files. Package data.table for big data.

library(readr)
library(tidyverse)

dat_abm <- read_csv("abm_short.csv", col_types = cols(casenum = col_character(), 
    year = col_character(), month = col_character(), 
    abm = col_character()))   # Because we created new project, we can use relative filepath.

# Click menu file> import data can conviently help you load data.


# 2. load dataset from online depository.

# depends on the online data format, typically you will put the url into the function for that format.


# 3. load dataset from package. install the package from github.
# remotes::install_github("higgi13425/medicaldata")


data(abm, package="medicaldata") #getting abm data from medicaldata package

dat_abm <- abm




# 4. connect to database and get data
#Use package to build connect and then SQL to get data.


# Sometimes we want to save the clean data for analysis.

# Save data as an object to a file
saveRDS(dat_abm, file = "dat_abm.rds")
# Read the rds data.
dat_abm <- readRDS(file = "dat_abm.rds")

# View(dat_abm) #view dataset in a seperate table.

# ??abm

# abm data is from medicaldata package, which is a simulated dataset of 1000 patients with acute bacterial meningitis (ABM). It contains 20 variables, including demographic information, clinical features, laboratory results, and outcomes.
```

## Data structure

```{r}
#| label: data-structure
#| eval: false
#| echo: true
# dimension
dim(dat_abm)

# detail structures.
str(dat_abm)

# variable names
colnames(dat_abm)

#change column names
colnames(dat_abm)[1] <- "id"

# print out the dataset can know a lot of info.
dat_abm
```


## 5 big verses from tidyverse

```{r}
#| label: tidyverse-verbs
#| eval: false
#| echo: true

#install.packages("tidyverse") #1st time require installing the package.


library(tidyverse) # typically we load all the required package at the beginning of the script, so we do not need to load them one by one later when we need them. Here just for demonstration, we will load tidyverse here.


# tibble is a modern reimagining of the data frame, providing a more user-friendly and efficient way to handle tabular data in R. It is part of the tidyverse collection of packages and offers several advantages over traditional data frames, such as better printing, enhanced subsetting, and improved handling of missing values. And won't change character into factor by default, which is a common source of confusion for new R users.

dat_vet <- tibble(
  animal_id = 1:6,
  species = c("Canine", "Feline", "Canine", "Equine", "Feline", "Canine"),
  weight_kg = c(25, 4, 32, 500, 5, 28),
  white_cell_count = c(10.5, 8.2, NA, 7.1, 9.4, 11.2),
  clinic_region = c("North", "South", "North", "West", "South", "North")
)

#1. filter(): Selecting Rows based on Conditions

dat_vet |>  
  filter(species == "Canine", weight_kg > 25)


#2. select(): Choosing Columns by names. you  can use helper function to select columns by pattern, such as starts_with(), ends_with(), contains() and matches().

dat_vet |>  
  select(animal_id, white_cell_count)


#3. mutate(): Creating New Variables/columns by transforming existing ones
dat_vet |> 
  mutate(weight_lb = weight_kg * 2.204,
         is_high_count = white_cell_count > 10)


#4. summarise(): Collapsing Data to a Value, usally used with group_by() to get summary statistics by group.

dat_vet  |>  
  summarise(med_weight = median(weight_kg, na.rm = TRUE),
            total_n = n())

dat_vet  |>  
  group_by(species) |>
  summarise(med_weight = median(weight_kg, na.rm = TRUE),
            total_n = n())


#5. arrange(): Ordering Data acording to values of a variable, default is ascending order, use desc() for descending order.
dat_vet |>  
  arrange(desc(weight_kg))


```



## Missing data

```{r}
#| label: check-missing-data
#| eval: false
#| echo: true

# some base funtions

is.na(dat_abm) # return TRUE for na position.

sum(is.na(dat_abm)) # TRUE is treated as 1 and FALSE is treated as 0, so sum(is.na(dat_abm)) gives the total number of missing values in the dataset.

#which(is.na(dat_abm)) #this is usually as condition to subset data.

dat_abm[which(is.na(dat_abm$race)),] #shows which rows has race having missing data.

# It is often useful to know how much is missing for each variable.

sum(is.na(dat_abm$race))/length(dat_abm$race)  

# It is fine we only have a few variables. If we have many variable, calculate this way sounds tedious.


# 1. apply function way

colSums(is.na(dat_abm))/nrow(dat_abm) # little better


# colMeans(is.na(dat_abm))
# apply(dat_abm, 2,function(x) sum(is.na(x))/length(x))  #basically a apply() function to loop over each variable


# 2. tidyverse way, much more convenient and easier to read.

dat_abm |> 
  summarise(across(everything(), ~ mean(is.na(.))))


```



## Univariate descriptive stats (Distribution)

### Continous variable

#### Central tendency

```{r}
#| label: central-tendency
#| eval: false
#| echo: true

# Mean

mean(dat_abm$age, na.rm=TRUE)  # note that a lot of function return unexpected results if you not pay attention to na.

# not fun to do one variable by one variable.

# 1. apply function way

colMeans(dat_abm[sapply(dat_abm,is.numeric)], na.rm=TRUE) # little better

#apply(dat_abm[sapply(dat_abm,is.numeric)], 2,mean,na.rm=TRUE)

# 2. tidyverse way

dat_abm |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))


# Median

# 1. apply function way
# no colMedians in base. There are packages however we will apply.
apply(dat_abm[sapply(dat_abm,is.numeric)], 2,median,na.rm=TRUE)
apply(dat_abm[sapply(dat_abm,is.numeric)], 2,quantile,probs=0.5,na.rm=TRUE)


# 2. tidyverse way

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ median(.x, na.rm = TRUE)))

# 1st quartile, 3rd quartile

# 1. apply function way
apply(dat_abm[sapply(dat_abm,is.numeric)], 2,quantile,probs=c(0.25,0.75),na.rm=TRUE)

# 2. tidyverse way

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  reframe(across(where(is.numeric), ~ quantile(.x, probs=c(0.25,0.75),na.rm = TRUE)))

# dat_abm %>%
#   summarise(across(where(is.numeric), list(first_quartile= ~ quantile(.x, probs=0.25,na.rm = TRUE),third_quartile= ~ quantile(.x, probs=0.75,na.rm = TRUE))))  # need manipulation to get into the desired form.

# minimum

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ min(.x,na.rm = TRUE)))

#maximum


dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ max(.x,na.rm = TRUE)))


```

#### Dispersion 

```{r}
#| label: dispersion
#| eval: false
#| echo: true
# standard deviation or variance
# this is the sample sd and variance, denominator is n-1
dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ sd(.x, na.rm = TRUE)))

# dat_abm %>%
#   summarise(across(where(is.numeric), ~ var(.x, na.rm = TRUE)))


# range  same information of min and max
dat_abm |>
  select(!where(haven::is.labelled)) |>
  reframe(across(where(is.numeric), ~ range(.x, na.rm = TRUE)))


# IQR

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ IQR(.x, na.rm = TRUE)))

# coefficient of variation


dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ (sd(.x, na.rm = TRUE))/(mean(.x, na.rm = TRUE))))


```


#### visualization

```{r}
#| label: univariate-visualization
#| eval: true
#| echo: true

#install.packages("ggplot2") #1st time require installing the package.

library(ggplot2)

# histogram

## base function
hist(dat_abm$blood_wbc, breaks =c(seq(0, 80, 2)), prob=TRUE)
lines(density(dat_abm$blood_wbc, bw=2,na.rm=T))
rug(dat_abm$blood_wbc)


# ggplot way
# ggplot is a great visualization tool R is famous for. Itself worth a course/tutorial. please learn as we needed.

ggplot(dat_abm, aes(x=blood_wbc)) + 
  geom_histogram(breaks =c(seq(0, 80, 2)))+
  # scale_x_log10()+
  xlab("WBC")+
  ylab("count")+
  ggtitle("Histogram for WBC variable at Bin size = 3")+
  theme_minimal()

  

```


### Categorical variable

#### Frequency and proportion

```{r}
#| label: categorical-frequency
#| eval: false
#| echo: true
# frequency or proportion for each category for each variable.

# because each variable has different categories we need to summarize one by one
dat_abm |>  
  select_if(is.character) 

table(dat_abm$race)
prop.table(table(dat_abm$race))

# Tidyverse way

dat_abm |> 
  group_by(race) |> 
  summarise(number=n())

dat_abm |> 
  group_by(race) %>%  
  summarise(number=n(), proportion=number/nrow(dat_abm))


```
### Visualization

```{r}
#| label: categorical-visualization
#| eval: true
#| echo: true

#barplot
ggplot(dat_abm, aes(x=haven::as_factor(race), fill=haven::as_factor(race))) +
  geom_bar() +
  xlab("Race") +
  ylab("Count") +
  ggtitle("Barplot for Race") +
  theme_classic()


```

### Super easy way to do univariate descriptive analysis.

```{r}
#| label: hmisc-univariate
#| eval: true
#| echo: true

library(Hmisc)

d <- dat_abm  |>  
  select(!where(haven::is.labelled)) |>
 Hmisc::describe()

Hmisc::html(d)
```



## Multivariate descriptive stats (Distribution)

Choose a variable as structural variable. They are usaully some categorical variable help us to stratify.

### Continous variable

#### Central tendency

```{r}
#| label: multivariate-central-tendency
#| eval: true
#| echo: true
# We can stratify by race

dat_abm |> 
  group_by(race) |>  
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))


```

#### Dispersion 

```{r}
#| label: multivariate-dispersion
#| eval: true
#| echo: true
# again stratify by race
dat_abm |> 
  group_by(race) |> 
  summarise(across(where(is.numeric), ~ sd(.x, na.rm = TRUE)))
```


#### visualization

```{r}
#| label: multivariate-visualization
#| eval: true
#| echo: true
# Since structural variable is categorical, we can do boxplot, or something similar like violin plot

dat_clean <- dat_abm |> 
  haven::zap_labels()# remove labels, which is a common source of confusion for new R users, and won't change character into factor by default, which is a common source of confusion for new R users.



# boxplot
dat_clean |> 
  ggplot( aes(x=race, y=blood_wbc, fill=race)) + # I add not delete NA normally won't include
    geom_boxplot()+
    geom_jitter(color="black", size=0.4, alpha=0.9) + # adding datapoints.
    ggtitle("A boxplot for WBC by race")

# violinplot
dat_clean |> 
  ggplot( aes(x=race, y=blood_wbc, fill=race)) + # I add not delete NA normally won't include
    geom_violin()+
    geom_jitter(color="black", size=0.4, alpha=0.9) + # adding datapoints.
    ggtitle("A violinplot for WBC by race")

```


### Categorical variable

#### Frequency and proportion

```{r}
#| label: contingency-table
#| eval: true
#| echo: true
# Stratify categorical variables will involve crosstable.

## 2 by 2 contigency table

table(dat_abm$race,dat_abm$sex)

# cell proportion
prop.table(table(dat_abm$race,dat_abm$sex))

# margin proportion
prop.table(table(dat_abm$race,dat_abm$sex),margin = 1) #margin = 1 specify row wise sum to 1
prop.table(table(dat_abm$race,dat_abm$sex),margin = 2) #margin = 2 specify column wise sum to 1


```
#### Visualization

```{r}
#| label: mosaic-plot
#| eval: true
#| echo: true
# remotes::install_github("haleyjeppson/ggmosaic") # get package from github.
library(ggmosaic)

dat_clean |> 
  mutate(race=as.factor(race),sex=as.factor(sex)) |> 
  filter(!is.na(race),!is.na(sex)) |>  ggplot() + 
  geom_mosaic(aes(x=product(race, sex), fill = race))


#base R
mosaicplot(table(dat_clean$race, dat_clean$abm))


```

### Super easy way to do multivariate descriptive analysis (table).

```{r}
#| label: hmisc-multivariate
#| eval: true
#| echo: true
s <-  Hmisc::summaryM(blood_gluc+blood_neut_pct+blood_wbc+age~ abm, data=dat_abm, overall=TRUE, test = TRUE )



Hmisc::html(
  s,
  caption = 'Multivariate descriptive analysis by case/control',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  pctdig =3,
  prmsd = TRUE,
  brmsd = TRUE,
  #msdsize = mu$smaller2
)
```



## Interrelationship between variable.

### Correlation

```{r}
#| label: correlation
#| eval: true
#| echo: true

# A lot of time we want to check correlation between continuous variable, especially for predictors. 

# In ABM dataset we assume all the continous variables are candidate preditors. Normally, we will decide which ones are candidate variables if too many of them.

# correlation matrix

cor(dat_abm %>% select_if(is.numeric), method = c("spearman"),use= "pairwise.complete.obs")


```


### Visualiziton

```{r warning=FALSE}
#| label: ggpairs-visualization
#| eval: true
#| echo: true
library(GGally)

# Let's just show a few
dat_abm |>  
  select(blood_wbc,csf_prot,blood_gluc) %>% 
ggpairs( title="correlogram with ggpairs()") 

```


## A lot of more to explore!!!



```{r}
#| label: session-info
#| eval: true
#| echo: true

# At the end of each script, list out all session information includes R and packages versions. 

sessionInfo()
```