---
title: "Intro to R for Researchers"
author: 
  - name: Sichao Wang
    degrees: PhD
    email: wangsic1@msu.edu
    affiliations: 
      - name: Michigan State University, Center for Statistical Training and Consulting
date: "`r format(Sys.time(), '%Y-%m-%d, %H:%M:%S %Z')`"
title-block-banner: "#01665e"
format: 
  html: 
    toc: true 
    toc-depth: 4
    toc-location: left 
    number-sections: true
    number-depth: 4
    code-tools: false 
    embed-resources: true 
    theme: default 
execute: 
  eval: true 
  echo: false #fenced
  output: true 
  message: false 
  warning: false 
  error: false 
  include: true 
knitr:
  opts_chunk:
    out.width: "150%"
bibliography: references.bib
editor:
  markdown:
    wrap: sentence
---

# Project title: Intro to R for Researchers

Description: "This is an introduction to R for researchers. It covers the basics of R programming, data manipulation, and data visualization. It also includes some examples of how to use R for data analysis."




# R built-in Help facility
## Help functions

```{r}

help.start()  # web-based r help pages.

help(mean)    # bring up the help for the specific function in the help panel.

?mean         # alternative way of help(mean)


methods(class="data.frame") #showing generic function for data.frame class

methods(plot) #what class generic function takes

getS3method("print", "data.frame") #method used for data frames
getS3method("print", "lm") #method used for linear model results

plot #  UseMethod("plot") means generic

getAnywhere("plot.acf")

version #check current R version

```


## Demostration

```{r}
demo()

demo(package = .packages(all.available = TRUE))

demo(lm.glm)
# demo(notebook,package = "knitr")

help(plotmath)
example(plotmath)
demo(plotmath)
```


# Key concepts

## Objects

### Types of object

1. atomic *vector*

```{r}
z <- rnorm(100)
```


2. *matrices* or more generally *arrays* are multi-dimensional generalizations of vectors.


```{r}
# An array can be considered as a multiply subscripted collection of data entries, for example numeric
dim(z) <- c(2,5,10) #2 by 5 by 10 array
```

3. *factors* provides compact ways to handle categorical data

order or unorder

```{r}
# unordered
fruits <- c("blueberry","strawberry","apple","pineapple","watermelon")
fruits_f <- factor(fruits)

levels(fruits_f)

levels(fruits_f) <- c("blueberry","strawberry","apple","pineapple","watermelon")

# Ordered

fruits_order <- factor(fruits,ordered = TRUE)




levels(fruits_order) <- c("blueberry","strawberry","apple","pineapple","watermelon")

fruits_order

unclass(fruits_order) #removes the "class" attribute from an object.

as.numeric(fruits_order)
```


4. *lists* are a general form of vector in which the various elements need not be of the same type, and are often themselves vectors or lists.


```{r}
Lst <- list(student="Joe", subject="Math", no.exam=3, scores=c(85,89,93))

Lst[[2]]
Lst$subject
Lst[["subject"]]

Lst[[4]][1]
Lst$scores[1]
Lst[["scores"]][1]

```


5. *data frames* are matrix-like structures (2 dimension), in which the columns can be of different types. 

A data frame may for many purposes be regarded as a matrix with columns possibly of differing modes and attributes. It may be displayed in matrix form, and its rows and columns extracted using matrix indexing conventions

```{r}
fruits <- data.frame(names=c("blueberry","apple","pineapple"),
                     color=c("blue","red","pineapple"),
                     weigh_lb=c(0.05,0.5,5))
```


5. *functions* are themselves objects in R which can be stored in the project’s workspace. This provides a simple and convenient way to extend R. 

All objects in R have a class, reported by the function class. For simple vectors this is just the mode, for example "numeric", "logical", "character" or "list", but "matrix", "array", "factor" and "data.frame" are other possible values


### Proporties of object


```{r}


mode(z) # the type of atomic component

length(z) # how many of the basic component


attr(z, "dim") <- c(10,10)

attributes(z)

```

### Demo for object 

```{r}
x <- rnorm(100) 
y <- rnorm(x)

plot(x, y)

ls() # list all the objects lives in the environment.
rm(x, y) # delete objects from environment.
```



```{r}
x <- 1:20
w <- 1 + sqrt(x)/2
dummy <- data.frame(x=x, y= x + rnorm(x)*w)
dummy
fm <- lm(y ~ x, data=dummy) 
summary(fm)

fm1 <- lm(y ~ x, data=dummy, weight=1/w^2)
summary(fm1)
attach(dummy) #Make the columns in the data frame visible as variables.

lrf <- lowess(x, y)
plot(x, y)
lines(x, lrf$y)
abline(0, 1, lty=3)
abline(coef(fm))
abline(coef(fm1), col = "red")
detach() #Remove data frame from the search path.

# Residuals vs. fitted plot

plot(fitted(fm), 
     resid(fm), 
     xlab="Fitted values", 
     ylab="Residuals", 
     main="Residuals vs Fitted")

# QQ plot for residuals

qqnorm(resid(fm), main="Residuals Rankit Plot")
qqline(resid(fm))
rm(fm, fm1, lrf, x, dummy)

```


## Data types

### numeric

```{r}
a<-2
b <- a
class(a)

# numeric vector
vec1<-c(1,2,3,4,5,6)
vec2<-c(1,2)

class(vec1)

vec1 + vec1 

vec1 / vec2  # notice that shorter vector will be recycled for operation.

sqrt(vec2)

n <- 10
1:n-1
1:(n-1) # notice the mathematical operation order, () take precedence.

# useful way to generate sequence. pretty useful to create index sequence in for loop, or plot ticks labels.

seq(1,30)
seq(from=1, to=30)
seq(to=30, from=1)

```
There are some common mathematical operator we can apply to numeric, such as  +, -, *, / and ^ . Also some of functions carry mathematical operation, such as log, exp, sin, cos, tan, sqrt.

### Character string

```{r}
?Quotes

char1 <- "Epi"  

class(char1)

char2 <- 'Epi'

char3 <- '"Epi"'


```

### logical

```{r}
logical1 <- TRUE
logical2 <- FALSE

class(logical2)

logical1+logical2


# A logical vector.

logical2 <- c(TRUE, TRUE,FALSE)

#as.something to help you manipulate types.

as.numeric(logical1) #coerced into numeric vectors, FALSE becoming 0 and TRUE becoming 1.


as.character(logical1)
```
There are also data type as integer, complex and raw, which are not very commonly used. Be careful about as.something, it might have unexpected result.


### Missing data

In general any operation on an NA becomes an NA, x == NA is quite different from is.na(x).
Not a Number, NaN, like denomitor is 0. In summary, is.na(xx) is TRUE both for NA and NaN values. To differentiate these, is.nan(xx) is only TRUE for NaNs.

```{r}
z <- c(1:3,NA); ind <- is.na(z)

z == NA
```


## Indexing(subsetting)


```{r}
x <- c(rnorm(10))

# by index number
x[1:3] # A vector of positive integral quantities select by position
x[-3]  # A vector of negative integral quantities deselect by position

# by conditional
y <- x[!is.na(x)&x>0] 

# by name
fruit <- c(5, 10, 1, 20) 
names(fruit) <- c("orange", "banana", "apple", "peach") 
lunch <- fruit[c("apple","orange")]

#index help you to manipulate vector such as

x[2] <- 44 # replace an element.
x[11] <- 100 # extend vector. 



```

Some conditional ooperator  <, <=, >, >=, == for exact equality and != for inequality
c1 & c2 is their intersection (“and”), c1 | c2 is their union (“or”), and !c1 is the negation of c1.

## Distribution

Prefix the name given here by ‘d’ for the density, ‘p’ for the CDF, ‘q’ for the quantile function and ‘r’ for simulation (r andom deviates)

```{r}
2*pt(-2.43, df = 13)
qf(0.01, 2, 7, lower.tail = FALSE)
```



# IDA: Descriptive data analysis

## Import data

There are a few ways to import/load dataset into your workspace.

1. load dataset from your local directory.
2. load dataset from online depository.
3. load dataset from package.
4. connect to database and get data


```{r}
# 1. load dataset from your local directory.
# Depends on the data format, you might need different function from different package to load data. Common package readxl for xls and xlsx files. Package data.table for big data.

library(readr)
library(tidyverse)

dat_abm <- read_csv("abm_short.csv", col_types = cols(casenum = col_character(), 
    year = col_character(), month = col_character(), 
    abm = col_character()))   # Because we created new project, we can use relative filepath.

# Click menu file> import data can conviently help you load data.


# 2. load dataset from online depository.

# depends on the online data format, typically you will put the url into the function for that format.


# 3. load dataset from package.
# remotes::install_github("higgi13425/medicaldata")


data(abm, package="medicaldata") #getting abm data from medicaldata package

dat_abm <- abm




# 4. connect to database and get data
#Use package to build connect and then SQL to get data.


# Sometimes we want to save the clean data for analysis.

# Save data as an object to a file
saveRDS(dat_abm, file = "dat_abm.rds")
# Read the rds data.
dat_abm <- readRDS(file = "dat_abm.rds")

# View(dat_abm) #view dataset in a seperate table.

# ??abm

```

## Data structure

```{r}

# dimension
dim(dat_abm)

# detail structures.
str(dat_abm)

# variable names
colnames(dat_abm)

#change column names
colnames(dat_abm)[1] <- "num"

# print out the dataset can know a lot of info.
dat_abm
```


## 5 big verses from tidyverse

```{r}
library(tidyverse)


dat_vet <- tibble(
  animal_id = 1:6,
  species = c("Canine", "Feline", "Canine", "Equine", "Feline", "Canine"),
  weight_kg = c(25, 4, 32, 500, 5, 28),
  white_cell_count = c(10.5, 8.2, NA, 7.1, 9.4, 11.2),
  clinic_region = c("North", "South", "North", "West", "South", "North")
)

#1. filter(): Selecting Rows

dat_vet |>  
  filter(species == "Canine", weight_kg > 25)


#2. select(): Choosing Columns
dat_vet |>  
  select(animal_id, white_cell_count)


#3. mutate(): Creating New Variables
dat_vet |> 
  mutate(weight_lb = weight_kg * 2.204,
         is_high_count = white_cell_count > 10)


#4. summarise(): Collapsing Data to a Value

dat_vet %>% 
  summarise(med_weight = median(weight_kg, na.rm = TRUE),
            total_n = n())

#5. arrange(): Ordering Data
dat_vet |>  
  arrange(desc(weight_kg))


```



## Missing data

```{r}
# some base funtions

is.na(dat_abm) # return TRUE for na

sum(is.na(dat_abm)) # remeber TRUE forced into 1, so we can get sum.

#which(is.na(dat_abm)) #this is usually as condition to subset data.

dat_abm[which(is.na(dat_abm$race)),] #shows which rows has race having missing data.

# It is often useful to know how much is missing for each variable.

sum(is.na(dat_abm$race))/length(dat_abm$race)  

# It is fine we only have a few variables. If we have many variable, calculate this way sounds tedious.


# 1. apply function way

colSums(is.na(dat_abm))/nrow(dat_abm) # little better
# colMeans(is.na(dat_abm))
# apply(dat_abm, 2,function(x) sum(is.na(x))/length(x))  #basically a apply() function to loop over each variable

# library(tidyverse)

# 2. tidyverse way

dat_abm |> 
  summarise(across(everything(), ~ mean(is.na(.))))


```



## Univariate descriptive stats (Distribution)

### Continous variable

#### Central tendency

```{r}
# Mean

mean(dat_abm$age, na.rm=TRUE)  # note that a lot of function return unexpected results if you not pay attention to na.

# not fun to do one variable by one variable.

# 1. apply function way

colMeans(dat_abm[sapply(dat_abm,is.numeric)], na.rm=TRUE) # little better

#apply(dat_abm[sapply(dat_abm,is.numeric)], 2,mean,na.rm=TRUE)

# 2. tidyverse way

dat_abm |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))


# Median

# 1. apply function way
# no colMedians in base. There are packages however we will apply.
apply(dat_abm[sapply(dat_abm,is.numeric)], 2,median,na.rm=TRUE)
apply(dat_abm[sapply(dat_abm,is.numeric)], 2,quantile,probs=0.5,na.rm=TRUE)


# 2. tidyverse way

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ median(.x, na.rm = TRUE)))

# 1st quartile, 3rd quartile

# 1. apply function way
apply(dat_abm[sapply(dat_abm,is.numeric)], 2,quantile,probs=c(0.25,0.75),na.rm=TRUE)

# 2. tidyverse way

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  reframe(across(where(is.numeric), ~ quantile(.x, probs=c(0.25,0.75),na.rm = TRUE)))

# dat_abm %>%
#   summarise(across(where(is.numeric), list(first_quartile= ~ quantile(.x, probs=0.25,na.rm = TRUE),third_quartile= ~ quantile(.x, probs=0.75,na.rm = TRUE))))  # need manipulation to get into the desired form.

# minimum

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ min(.x,na.rm = TRUE)))

#maximum


dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ max(.x,na.rm = TRUE)))


```

#### Dispersion 

```{r}

# standard deviation or variance
# this is the sample sd and variance, denominator is n-1
dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ sd(.x, na.rm = TRUE)))

# dat_abm %>%
#   summarise(across(where(is.numeric), ~ var(.x, na.rm = TRUE)))


# range  same information of min and max
dat_abm |>
  select(!where(haven::is.labelled)) |>
  reframe(across(where(is.numeric), ~ range(.x, na.rm = TRUE)))


# IQR

dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ IQR(.x, na.rm = TRUE)))

# coefficient of variation


dat_abm |> 
  select(!where(haven::is.labelled)) |>
  summarise(across(where(is.numeric), ~ (sd(.x, na.rm = TRUE))/(mean(.x, na.rm = TRUE))))


```


#### visualization

```{r}
library(ggplot2)

# histogram

## base function
hist(dat_abm$blood_wbc, breaks =c(seq(0, 80, 2)), prob=TRUE)
lines(density(dat_abm$blood_wbc, bw=2,na.rm=T))
rug(dat_abm$blood_wbc)


# ggplot way
# ggplot is a great visualization tool R is famous for. Itself worth a course/tutorial. please learn as we needed.

ggplot(dat_abm, aes(x=blood_wbc)) + 
  geom_histogram(breaks =c(seq(0, 80, 2)))+
  # scale_x_log10()+
  xlab("WBC")+
  ylab("count")+
  ggtitle("Histogram for WBC variable at Bin size = 3")+
  theme_minimal()

  

```


### Categorical variable

#### Frequency and proportion

```{r}

# frequency or proportion for each category for each variable.

# because each variable has different categories we need to summarize one by one
dat_abm |>  
  select_if(is.character) 

table(dat_abm$race)
prop.table(table(dat_abm$race))

# Tidy verse way

dat_abm |> 
  group_by(race) |> 
  summarise(number=n())

dat_abm |> 
  group_by(race) %>%  
  summarise(number=n(), proportion=number/nrow(dat_abm))


```
### Visualization

```{r}
#barplot

ggplot(dat_abm, aes(x=as.factor(race), fill=as.factor(race))) +  
  geom_bar( )
```

### Super easy way to do univariate descriptive analysis.

```{r}
library(Hmisc)



d <- dat_abm  |>  
  select(!where(haven::is.labelled)) |>
 Hmisc::describe()

Hmisc::html(d)
```



## Multivariate descriptive stats (Distribution)

Choose a variable as structural variable. They are usaully some categorical variable help us to stratify.

### Continous variable

#### Central tendency

```{r}

# We can stratify by race

dat_abm |> 
  group_by(race) |>  
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))


```

#### Dispersion 

```{r}

# again stratify by race
dat_abm |> 
  group_by(race) |> 
  summarise(across(where(is.numeric), ~ sd(.x, na.rm = TRUE)))
```


#### visualization

```{r}
# Since structural variable is categorical, we can do boxplot, or something similar like violin plot

dat_clean <- dat_abm |> 
  haven::zap_labels()



# boxplot
dat_clean |> 
  ggplot( aes(x=race, y=blood_wbc, fill=race)) + # I add not delete NA normally won't include
    geom_boxplot()+
    geom_jitter(color="black", size=0.4, alpha=0.9) + # adding datapoints.
    ggtitle("A boxplot for WBC by race")

# ciolin
dat_clean |> 
  ggplot( aes(x=race, y=blood_wbc, fill=race)) + # I add not delete NA normally won't include
    geom_violin()+
    geom_jitter(color="black", size=0.4, alpha=0.9) + # adding datapoints.
    ggtitle("A boxplot for WBC by race")

```


### Categorical variable

#### Frequency and proportion

```{r}

# Stratify categorical variables will involve crosstable.

## 2 by 2 contigency table

table(dat_abm$race,dat_abm$sex)

# cell proportion
prop.table(table(dat_abm$race,dat_abm$sex))

# margin proportion
prop.table(table(dat_abm$race,dat_abm$sex),margin = 1) #margin = 1 specify row wise sum to 1
prop.table(table(dat_abm$race,dat_abm$sex),margin = 2) #margin = 2 specify column wise sum to 1


```
#### Visualization

```{r}
# remotes::install_github("haleyjeppson/ggmosaic")
library(ggmosaic)

dat_clean |> 
  mutate(race=as.factor(race),sex=as.factor(sex)) |> 
  filter(!is.na(race),!is.na(sex)) |>  ggplot() + 
  geom_mosaic(aes(x=product(race, sex), fill = race))


#base R
mosaicplot(table(dat_clean$race, dat_clean$abm))


```

### Super easy way to do multivariate descriptive analysis (table).

```{r}
s <-  Hmisc::summaryM(blood_gluc+blood_neut_pct+blood_wbc+age~ abm, data=dat_abm, overall=TRUE, test = TRUE )



Hmisc::html(
  s,
  caption = 'Multivariate descriptive analysis by case/control',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  pctdig =3,
  prmsd = TRUE,
  brmsd = TRUE,
  #msdsize = mu$smaller2
)
```



## Interrelationship between variable.

### Correlation

```{r}
# A lot of time we want to check correlation between continuous variable, especially for predictors. 

# In ABM dataset we assume all the continous variables are candidate preditors. Normally, we will decide which ones are candidate variables if too many of them.

# correlation matrix

cor(dat_abm %>% select_if(is.numeric), method = c("spearman"),use= "pairwise.complete.obs")


```


### Visualiziton

```{r warning=FALSE}
library(GGally)

# Let's just show a few
dat_abm |>  
  select(blood_wbc,csf_prot,blood_gluc) %>% 
ggpairs( title="correlogram with ggpairs()") 

```


## A lot of more to explore!!!



